Effectful Semantics for Substructural SSA

### 2023-11-16 pldi24/paper.tex
Tap

We begin with some basic definitions:
\begin{definition}[Binoidal Category] 
  A \textbf{binoidal category} \(\mc{C}\) is a category equipped with a binary
  operation on objects \(\otimes: \mc{C} \times \mc{C} \to \mc{C}\) which is
  independently functorial in each argument (i.e., for each object \(A\), \(A
  \otimes -\) and \(- \otimes A\) are functors).
\end{definition}
Given \(f: \mc{C}(A, B)\), \(g: \mc{C}(A', B')\) in a binoidal category, we
write \(f \ltimes g = f \otimes A';B \otimes g\) and \(f \rtimes g = A \otimes
g;f \otimes B'\); we say \(f\) is \textbf{central} if, for all \(g\), \(f
\ltimes g = f \rtimes g\) and \(g \ltimes f = g \rtimes f\), in which case we
simply write \(f \otimes g\), \(g \otimes f\) respectively.
\begin{definition}[Premonoidal Category] 
  A \textbf{premonoidal category} is a binoidal category with an identity object
  \(I: \mc{C}\) such that
  \begin{itemize}
    \item There exist central natural isomorphisms \(\lambda_A: I \otimes A
    \simeq A\) (the \textbf{left unitor}), \(\rho_A: A \otimes I \simeq A\) (the
    \textbf{right unitor})
    \item There exists a central isomorphism \(\alpha_{A, B, C}: (A \otimes B)
    \otimes C \simeq A \otimes (B \otimes C)\) (the \textbf{associator})
    \item Satisfying the \textbf{pentagon law}: \todo{this}
    \item Satisfying the \textbf{triangle law}: \todo{this}
  \end{itemize}
  We say a premonoidal category is \textbf{symmetric} if it is equipped with a
  central natural isomorphism \(\sigma_{A, B}: A \otimes B \simeq B \otimes A\)
  (the \textbf{symmetry}) satisfying the \textbf{hexagon law} \todo{this}. We
  say a premonoidal category is \textbf{monoidal} if every morphism is central.
\end{definition}
One issue with the notion of a premonoidal category is that it is unclear what
exactly a premonoidal functor should be: if we want them to compose, we would
like to preserve central morphisms, but trying to preserve \textit{all} central
morphisms is too strict a notion. We want, in essence, to be able to distinguish
morphisms that are truly "pure" from those which simply happen to be central. To
do that, we introduce the notion of an \textit{effectful category} as follows:
\begin{definition}[Effectful Category] 
  An \textbf{effectful category} \(\mc{V} \to \mc{C}\) is an identity on objects
  functor \(\upg{\cdot}{}: \mc{V} \to \mc{C}\) from a monoidal category
  \(\mc{V}\) to a premonoidal category \(\mc{C}\) which preserves all
  premonoidal structure (i.e. sends associators to associators, unitors to
  unitors, and tensor products to tensor products). It is \textbf{symmetric} if
  both \(\mc{V}\) and \(\mc{C}\) are, and \(\upg{\cdot}{}\) preserves the
  symmetry. A \textbf{Freyd category} is an effectful category where \(\mc{V}\)
  is Cartesian.
\end{definition}

\TODO{confusion with notion of pure in truly "pure" above}

\TODO{text on Freyd categories}

\begin{definition}[(Symmetric) Premonoidal Subcategory]
  Given a (symmetric) premonoidal category \(\mc{C}\), we say a wide subcategory
  \(\mc{S} \subseteq \mc{C}\) is \textbf{(symmetric) premonoidal} if it is
  closed under whiskering and contains all associators, unitors, and symmetries.
  We say a subcategory is \textbf{monoidal} if, for all \(f: \mc{S}(A, B)\),
  \(g: \mc{S}(A', B')\), \(f \ltimes g = f \rtimes g\); we say it is
  \textbf{central} if this generalizes to the case where only one of \(f, g\) is
  in \(\mc{S}\) (i.e., if all morphisms in \(\mc{S}\) are central).
\end{definition} 
Note that the set of symmetric premonoidal subcategories, monoidal
subcategories, and central subcategories of a category \(\mc{C}\) are all closed
under intersections and therefore, in particular, form a complete
meet-semilattice with shared bottom element \(\ms{Skeleton}(\mc{C})\), where
\(\ms{Skeleton}(C)\) is the smallest possible symmetric premonodial subcategory.
The largest symmetric premonoidal subcategory is of course \(\mc{C}\), while the
largest central subcategory is the center of \(\mc{C}\), \(Z(\mb{C})\); the
existence of a top-element means that in these cases we may induce the structure
of a complete lattice in the usual manner.

% Should we go down this road?
% \begin{definition}[Strictly Effectful Category] \
%   \TODO{this}
% \end{definition}

\begin{definition}[Substructural (Pre)monoidal Category]
  A \textbf{substructural (pre)monoidal category} is a tuple \((\mc{C}, \mc{A},
  \mc{R})\) where:
  \begin{itemize}
    \item \(\mc{C}\) is a symmetric premonoidal category equipped with wide
    symmetric premonoidal subcategories \(\mc{C}^{\ms{a}}, \mc{C}^{\ms{r}}\).
    We define, for \(q \subseteq \{\ms{a}, \ms{r}\}\), \(\mc{C}^q = \bigcap_{\ell \in q}\mc{C}^\ell\), where \(\mc{C}^\varnothing = \mc{C}\)
    \item \(\mc{A} \subseteq |\mc{C}|\) is a set of \textbf{affine} objects
    closed under tensor products, such that
    \begin{itemize}
      \item \(\mb{1} \in \mc{A}\)
      \item Every \(A\) in \(\mc{A}\) is equipped with a central morphism
      \(\ms{drop}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, \mb{1})\)
      \item \(\forall A, B \in \mc{A}, \ms{drop}(A \otimes B) = \ms{drop}(A) \otimes \ms{drop}(B);\lambda_I\)
      \item \(\forall A, B \in \mc{A}, \forall f: \mc{C}^{\ms{a}}(A, B), f;\ms{drop}(B) = \ms{drop}(A)\)
    \end{itemize}
    \item \(\mc{R} \subseteq |\mc{C}|\) is a set of \textbf{relevant} objects
    closed under tensor products, such that
    \begin{itemize}
      \item \(\mb{1} \in \mc{R}\)
      \item Every \(A\) in \(\mc{R}\) is equipped with a central morphism
      \(\ms{split}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, A \otimes A)\)
      \item \(\forall A, B \in \mc{R}, \ms{split}(A \otimes B) = \ms{split}(A) \otimes \ms{split}(B);\alpha;A \otimes \sigma_{A, B} \otimes B;\alpha\)
      \item Associativity: \(\forall A \in \mc{R}, \ms{split}(A);\ms{split}(A) \otimes A;\alpha = \ms{split}(A);A \otimes \ms{split}(A)\)
      \item Commutativity: \(\forall A \in \mc{R}, \ms{split}(A);\sigma = \ms{split}(A)\)
      \item Comonoid: \(\forall A \in \mc{A} \cap \mc{R}, \ms{split}(A);A \otimes \ms{drop}(A) = \rho_A^{-1} \quad \ms{split}(A);\ms{drop}(A) \otimes A = \lambda_A^{-1}\)
      \item \(\forall A, B \in \mc{R}, \forall f: \mc{C}^{\ms{r}}(A, B), f;\ms{split}(B) = \ms{split}(A);f \ltimes f = \ms{split}(A);f\rtimes f\)
    \end{itemize}
  \end{itemize}
  We call the morphisms in \(\mc{C}^{\ms{a}}\) \textbf{affine}, the
  morphisms in \(\mc{C}^{\ms{r}}\) \textbf{relevant}, and the morphisms in
  \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) \textbf{pure} or \textbf{intuitionistic}. We
  call the morphisms in \(\mc{C}^\varnothing\) but not necessarily in any of the
  other \(\mc{C}^q\) \textbf{linear}. We call a substructural \textit{monoidal}
  category a \textbf{substructural Cartesian category} if \(\mc{A} = \mc{R} =
  |\mc{C}|\), in which case \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) is Cartesian.
\end{definition}
A simple example of a substructural Cartesian category is the category of
relations \(\ms{Rel}\), viewed as the Kleisli category of the power-set monad
with morphisms of the form \(\ms{Rel}(A, B) = A \to \mc{P}(B)\). In particular,
we can define:
\begin{equation}
  \begin{aligned}
    \ms{Rel}^{\ms{a}}(A, B) 
    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| \geq 1\} 
    &&= A \to \mc{P}^+(B) \\
    \ms{Rel}^{\ms{r}}(A, B) 
    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| \leq 1\} 
    &&= A \to \ms{Option}(B) \\
    \implies \ms{Rel}^{\{\ms{a}, \ms{r}\}}(A, B) 
    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| = 1\} 
    &&= A \to B
  \end{aligned}
\end{equation}
Here, \(\ms{split}(A)\) is just the copy operator \(a \mapsto (a, a)\) and
\(\ms{drop}(A)\) the deletion operator \(a \mapsto ()\); it is quite trivial to
prove that these lie in \(\ms{Rel}^{\{\ms{a}, \ms{r}\}}\), being functions, and
that they satisfy the axioms of a commutative comonoid, as desired.

Note that we cannot simply have \(\ms{Rel}^{\ms{a}} = \ms{Rel}\), since, for
any morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A, R(a) =
\varnothing\), we have
\begin{equation}
  (R;\ms{drop}(B))(a) = \varnothing \neq \ms{drop}(A)(a) = \{()\}
\end{equation}
We will see that this corresponds to the fact that the following programs are
not semantically equal
\begin{equation}
  \ms{let}\;x = \ms{loop}(); e \not\simeq e \qquad \text{where} \qquad x \notin \ms{fv}(e)
\end{equation}
Similarly, we cannot simply have \(\ms{Rel}^{\ms{r}} = \ms{Rel}\), since,
for any morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A, |R(a)|
\supseteq \{b_0, b_1\}\) where \(b_0 \neq b_1\), we have that
\begin{equation}
  (R;\ms{split}(B))(a) = \{(b, b) \mid b \in R(a)\} \neq (\ms{split}(A);R)(a) = R(a) \times R(a) 
\end{equation}
since only the left-hand side contains the pair \((b_0, b_1)\), this
corresponding to the fact that the following programs are not semantically equal
\begin{equation}
  \ms{let}\;x = \ms{rand}(); (x, x) \not\simeq (\ms{rand}(), \ms{rand}())
\end{equation}
To get a substructural premonoidal category, we can apply the state monad
transformer to the power set monad to obtain the Kleisli category with morphisms
of the form \(\ms{Set}_{\ms{St}\;\mc{P}\;S}(A, B) = A \to S \to \mc{P}(B \times
S)\). Once again, it is clear that all objects have split and drop morphisms
inherited from \(\ms{Set}\), and therefore are affine and relevant

\begin{definition}[Susbtructural Effectful Category]
  A \textbf{substructural effectful category} \(\mc{C}\) consists of
  \begin{itemize}
    \item A substructural monoidal category \(\mc{C}_1\); we call the morphisms
    in this category \textbf{central} or \textbf{linear}
    \item A substructural premonoidal category \(\mc{C}_0\) sharing the same
    affine and relevant objects as \(\mc{C}_1\)
    \item An identity-on-objects functor \(\upg{\cdot}{}: \mc{C}_1^\varnothing
    \to \mc{C}_0^\varnothing\) preserving all substructural premonoidal
    structure, i.e., preserves all symmetries, associators, unitors, drops, and
    splits and sends \(\mc{C}_1^q\) to \(\mc{C}_0^q\). In particular,
    restricting \(\upg{\cdot}{}\) appropriately makes all \(\mc{C}_1^q \to
    \mc{C}_0^q\) into effectful categories.
  \end{itemize}
  We define \(\upg{f}{0}\) to be \(\upg{f}{}\) for morphisms in \(\mc{C}_1^q\)
  and simply \(f\) for morphisms in \(\mc{C}_0\), and \(\upg{f}{1}\) to simply
  be the identity functor on \(\mc{C}_1^q\).

  Morphisms in \(\mc{C}_p^{\ms{r}}\) are called \textbf{copyable}, morphisms
  in \(\mc{C}_p^{\ms{a}}\) are called \textbf{discardable}, while morphisms
  in \(\mc{C}_1^{\ms{a}}, \mc{C}_1^{\ms{r}}, \mc{C}_1^{\{\ms{a},
  \ms{r}\}}\) specifically are called \textbf{affine}, \textbf{relevant}, and
  \textbf{intuitionistic} or \textbf{pure} respectively.
\end{definition}

We can obtain an example of a substructural effectful category quite easily by,
considering the power-set monad above, applying the state transformer to yield
the Kliesli category with morphisms of the form \(\mc{C}_0(A, B) = A \to S \to
\mc{P}(A \times S)\). Letting \(\mc{C}_1^q = \ms{Rel}^q\) and defining
\(
  \upg{R}{} = \lambda a, s. R(a) \times \{s\}
\), we can verify that this satisfies the axioms of a substructural effectful category.
Note that here the \textit{central} morphisms are precisely those which do not
read from or write to the state, and therefore which commute with all other
morphisms.

@@

\TODO{We know \textit{what} these semantics are, unlike below, but maybe clarify
\textit{why}}

@@

Now that we have a semantics for \isotopessa programs, we can show
that it matches the syntax by showing that the syntactic and semantic
notions of substitution coincide.

\todo{Move weakening and upgrade lemmas to appendix}
% Since we're
% approaching things from a type-theoretic perspective, we begin by addressing the
% semantic versions of the standard syntactic metatheorems of weakening and
% substitution. 
% 
% In particular, semantic weakening can be stated in the expected
% manner: if we have an expression or block which is well typed in both \(\Gamma\)
% and \(\Delta\), where \(\cwk{\Gamma}{\Delta}\), then evaluating it in \(\Gamma\)
% is the same as first strengthening \(\Gamma\) to \(\Delta\), dropping unused
% affine variables, and then evaluating it in \(\Delta\).
% \begin{lemma}[Semantic Weakening] 
%   If \(\cwk{\Gamma}{\Delta}\), then 
%   \begin{itemize}
%     \item For all \(\hasty{\Delta}{p}{a}{A}{q}\),
%     \[
%       \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
%       ; \dnt{\hasty{\Delta}{p}{a}{A}{q}}
%       = \dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
%     \]
%     \item For all \(\haslb{\Delta}{t}{\ms{L}}\),
%     \[
%       \upg{\dnt{\cwk{\Gamma}{\Delta}}}{p}
%       ; \dnt{\haslb{\Delta}{t}{\ms{L}}}
%       = \dnt{\haslb{\Gamma}{t}{\ms{L}}}  
%     \]
%   \end{itemize}
% 
%   Similarly, if \(\lwk{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma}{t}{\ms{L}}\), then
%   \[
%     \dnt{\haslb{\Gamma}{t}{\ms{L}}}
%     ; \dnt{\lwk{\ms{L}}{\ms{K}}}
%     = \dnt{\haslb{\Gamma}{t}{\ms{K}}}
%   \]
% \end{lemma}
% 
% \TODO{some text about upgrade}
% 
% \begin{lemma}[Semantic Upgrade]
%   For all \(\Gamma, a, A\), if \(\hasty{\Gamma}{1}{a}{A}{q}\), then
%   \[\dnt{\hasty{\Gamma}{0}{a}{A}{q}} = \upg{\dnt{\hasty{\Gamma}{1}{a}{A}{q}}}{}\]
% \end{lemma}

\begin{figure}
  \begin{gather*}
    \boxed{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}: \mc{C}_1^\varnothing(\dnt{\Theta}, \dnt{\Gamma})} 
    \\
    \dnt{\issubst{\gamma}{\Theta}{\Gamma, \thyp{x}{A}{q}}}
      = \dnt{\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}};\dnt{\issubst{\gamma}{\Theta_\Gamma}{\Gamma}} \otimes \dnt{\hasty{\Theta_x}{1}{a}{A}{q}} \\
    \dnt{\issubst{\cdot}{\cdot}{\cdot}} = \ms{id}
    \\
    \boxed{\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}: \mc{C}_0^\varnothing(\dnt{\ms{L}}, \dnt{\ms{K}})} 
  \end{gather*}
  \caption{Semantics for \isotopessa substitutions}
  \Description{Semantics for isotope-SSA substitutions}
  \label{fig:ssa-subst-semantics}
\end{figure}

\TODO{some text about substitution semantics -- substitution = composition of env semantics with therm semantics}
\TODO{some text about semantic substitution -- it's a function that does the obvious thing}

>>>>

We begin with some basic definitions:
\begin{definition}[Binoidal Category] 
  A \textbf{binoidal category} \(\mc{C}\) is a category equipped with a binary
  operation on objects \(\otimes: \mc{C} \times \mc{C} \to \mc{C}\) which is
  independently functorial in each argument (i.e., for each object \(A\), \(A
  \otimes -\) and \(- \otimes A\) are functors).
\end{definition}
Given \(f: \mc{C}(A, B)\), \(g: \mc{C}(A', B')\) in a binoidal category, we
write \(f \ltimes g = f \otimes A';B \otimes g\) and \(f \rtimes g = A \otimes
g;f \otimes B'\); we say \(f\) is \textbf{central} if, for all \(g\), \(f
\ltimes g = f \rtimes g\) and \(g \ltimes f = g \rtimes f\), in which case we
simply write \(f \otimes g\), \(g \otimes f\) respectively.
\begin{definition}[Premonoidal Category] 
  A \textbf{premonoidal category} is a binoidal category with an identity object
  \(I: \mc{C}\) such that
  \begin{itemize}
    \item There exist central natural isomorphisms \(\lambda_A: I \otimes A
    \simeq A\) (the \textbf{left unitor}), \(\rho_A: A \otimes I \simeq A\) (the
    \textbf{right unitor})
    \item There exists a central isomorphism \(\alpha_{A, B, C}: (A \otimes B)
    \otimes C \simeq A \otimes (B \otimes C)\) (the \textbf{associator})
    \item Satisfying the \textbf{pentagon law}: \todo{this}
    \item Satisfying the \textbf{triangle law}: \todo{this}
  \end{itemize}
  We say a premonoidal category is \textbf{symmetric} if it is equipped with a
  central natural isomorphism \(\sigma_{A, B}: A \otimes B \simeq B \otimes A\)
  (the \textbf{symmetry}) satisfying the \textbf{hexagon law} \todo{this}. We
  say a premonoidal category is \textbf{monoidal} if every morphism is central.
\end{definition}
One issue with the notion of a premonoidal category is that it is unclear what
exactly a premonoidal functor should be: if we want them to compose, we would
like to preserve central morphisms, but trying to preserve \textit{all} central
morphisms is too strict a notion. We want, in essence, to be able to distinguish
morphisms that are truly "pure" from those which simply happen to be central. To
do that, we introduce the notion of an \textit{effectful category} as follows:
\begin{definition}[Effectful Category] 
  An \textbf{effectful category} \(\mc{V} \to \mc{C}\) is an identity on objects
  functor \(\upg{\cdot}{}: \mc{V} \to \mc{C}\) from a monoidal category
  \(\mc{V}\) to a premonoidal category \(\mc{C}\) which preserves all
  premonoidal structure (i.e. sends associators to associators, unitors to
  unitors, and tensor products to tensor products). It is \textbf{symmetric} if
  both \(\mc{V}\) and \(\mc{C}\) are, and \(\upg{\cdot}{}\) preserves the
  symmetry. A \textbf{Freyd category} is an effectful category where \(\mc{V}\)
  is Cartesian.
\end{definition}
Unfortunately, the notion of "purity" given by an effectful category only allows
us to perform relatively weak optimizations in the presence of effects. To
remedy this, we want to be able to make a finer distinction between morphisms
which are merely \textit{central} and those which are also \textit{relevant}
(i.e., copying the input and applying them twice is the same as applying them
once and copying the output) and/or \textit{affine} (i.e., applying them and
discarding the result is the same as never doing them at all). Doing so also
allows us to add the capability of \textit{distinguishing} between objects which
are affine, relevant, neither or both, enabling us to reason about substructural type systems categorically.

To do this, we begin by introducing the notion of a \textit{symmetric premonoidal subcategory}, as follows:
\begin{definition}[(Symmetric) Premonoidal Subcategory]
  Given a (symmetric) premonoidal category \(\mc{C}\), we say a wide subcategory
  \(\mc{S} \subseteq \mc{C}\) is \textbf{(symmetric) premonoidal} if it is
  closed under whiskering and contains all associators, unitors, and symmetries.
  We say a subcategory is \textbf{monoidal} if, for all \(f: \mc{S}(A, B)\),
  \(g: \mc{S}(A', B')\), \(f \ltimes g = f \rtimes g\); we say it is
  \textbf{central} if this generalizes to the case where only one of \(f, g\) is
  in \(\mc{S}\) (i.e., if all morphisms in \(\mc{S}\) are central).
\end{definition} 
Note that the set of symmetric premonoidal subcategories, monoidal
subcategories, and central subcategories of a category \(\mc{C}\) are all closed
under intersections and therefore, in particular, form a complete
meet-semilattice with shared bottom element \(\ms{Skeleton}(\mc{C})\), where
\(\ms{Skeleton}(C)\) is the smallest possible symmetric premonodial subcategory.
The largest symmetric premonoidal subcategory is of course \(\mc{C}\), while the
largest central subcategory is the center of \(\mc{C}\), \(Z(\mb{C})\); the
existence of a top-element means that in these cases we may induce the structure
of a complete lattice in the usual manner.

We can now define a \textit{substructural premonoidal category}, which is simply
a premonoidal category equipped with a distinguished set of relevant and affine
morphisms
\begin{definition}[Substructural (Pre)monoidal Category]
  A \textbf{substructural (pre)monoidal category} is a tuple \((\mc{C}, \mc{A},
  \mc{R})\) where:
  \begin{itemize}
    \item \(\mc{C}\) is a symmetric premonoidal category equipped with wide
    symmetric premonoidal subcategories \(\mc{C}^{\ms{a}}, \mc{C}^{\ms{r}}\).
    We define, for \(q \subseteq \{\ms{a}, \ms{r}\}\), \(\mc{C}^q = \bigcap_{\ell \in q}\mc{C}^\ell\), where \(\mc{C}^\varnothing = \mc{C}\)
    \item \(\mc{A} \subseteq |\mc{C}|\) is a set of \textbf{affine} objects
    closed under tensor products, such that
    \begin{itemize}
      \item \(\mb{1} \in \mc{A}\)
      \item Every \(A\) in \(\mc{A}\) is equipped with a central morphism
      \(\ms{drop}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, \mb{1})\)
      \item \(\forall A, B \in \mc{A}, \ms{drop}(A \otimes B) = \ms{drop}(A) \otimes \ms{drop}(B);\lambda_I\)
      \item \(\forall A, B \in \mc{A}, \forall f: \mc{C}^{\ms{a}}(A, B), f;\ms{drop}(B) = \ms{drop}(A)\)
    \end{itemize}
    \item \(\mc{R} \subseteq |\mc{C}|\) is a set of \textbf{relevant} objects
    closed under tensor products, such that
    \begin{itemize}
      \item \(\mb{1} \in \mc{R}\)
      \item Every \(A\) in \(\mc{R}\) is equipped with a central morphism
      \(\ms{split}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, A \otimes A)\)
      \item \(\forall A, B \in \mc{R}, \ms{split}(A \otimes B) = \ms{split}(A) \otimes \ms{split}(B);\alpha;A \otimes \sigma_{A, B} \otimes B;\alpha\)
      \item Associativity: \(\forall A \in \mc{R}, \ms{split}(A);\ms{split}(A) \otimes A;\alpha = \ms{split}(A);A \otimes \ms{split}(A)\)
      \item Commutativity: \(\forall A \in \mc{R}, \ms{split}(A);\sigma = \ms{split}(A)\)
      \item Comonoid: \(\forall A \in \mc{A} \cap \mc{R}, \ms{split}(A);A \otimes \ms{drop}(A) = \rho_A^{-1} \quad \ms{split}(A);\ms{drop}(A) \otimes A = \lambda_A^{-1}\)
      \item \(\forall A, B \in \mc{R}, \forall f: \mc{C}^{\ms{r}}(A, B), f;\ms{split}(B) = \ms{split}(A);f \ltimes f = \ms{split}(A);f\rtimes f\)
    \end{itemize}
  \end{itemize}
  We call the morphisms in \(\mc{C}^{\ms{a}}\) \textbf{affine}, the
  morphisms in \(\mc{C}^{\ms{r}}\) \textbf{relevant}, and the morphisms in
  \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) \textbf{pure} or \textbf{intuitionistic}. We
  call the morphisms in \(\mc{C}^\varnothing\) but not necessarily in any of the
  other \(\mc{C}^q\) \textbf{linear}. We call a substructural \textit{monoidal}
  category a \textbf{substructural Cartesian category} if \(\mc{A} = \mc{R} =
  |\mc{C}|\), in which case \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) is Cartesian.
\end{definition}
A simple example of a substructural Cartesian category is the category of
relations \(\ms{Rel}\), viewed as the Kleisli category of the power-set monad
with morphisms of the form \(\ms{Rel}(A, B) = A \to \mc{P}(B)\). In particular,
we can define:
\begin{equation}
  \begin{aligned}
    \ms{Rel}^{\ms{a}}(A, B) 
    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| \geq 1\} 
    &&= A \to \mc{P}^+(B) \\
    \ms{Rel}^{\ms{r}}(A, B) 
    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| \leq 1\} 
    &&= A \to \ms{Option}(B) \\
    \implies \ms{Rel}^{\{\ms{a}, \ms{r}\}}(A, B) 
    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| = 1\} 
    &&= A \to B
  \end{aligned}
\end{equation}
Here, \(\ms{split}(A)\) is just the copy operator \(a \mapsto (a, a)\) and
\(\ms{drop}(A)\) the deletion operator \(a \mapsto ()\); it is quite trivial to
prove that these lie in \(\ms{Rel}^{\{\ms{a}, \ms{r}\}}\), being functions, and
that they satisfy the axioms of a commutative comonoid, as desired.

Note that we cannot simply have \(\ms{Rel}^{\ms{a}} = \ms{Rel}\), since, for
any morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A, R(a) =
\varnothing\), we have
\begin{equation}
  (R;\ms{drop}(B))(a) = \varnothing \neq \ms{drop}(A)(a) = \{()\}
\end{equation}
We will see that this corresponds to the fact that the following programs are
not semantically equal
\begin{equation}
  \ms{let}\;x = \ms{loop}(); e \not\simeq e \qquad \text{where} \qquad x \notin \ms{fv}(e)
\end{equation}
Similarly, we cannot simply have \(\ms{Rel}^{\ms{r}} = \ms{Rel}\), since,
for any morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A, |R(a)|
\supseteq \{b_0, b_1\}\) where \(b_0 \neq b_1\), we have that
\begin{equation}
  (R;\ms{split}(B))(a) = \{(b, b) \mid b \in R(a)\} \neq (\ms{split}(A);R)(a) = R(a) \times R(a) 
\end{equation}
since only the left-hand side contains the pair \((b_0, b_1)\), this
corresponding to the fact that the following programs are not semantically equal
\begin{equation}
  \ms{let}\;x = \ms{rand}(); (x, x) \not\simeq (\ms{rand}(), \ms{rand}())
\end{equation}
To get a substructural premonoidal category, we can apply the state monad
transformer to the power set monad to obtain the Kleisli category with morphisms
of the form \(\ms{Set}_{\ms{St}\;\mc{P}\;S}(A, B) = A \to S \to \mc{P}(B \times
S)\). Once again, it is clear that all objects have split and drop morphisms
inherited from \(\ms{Set}\), and therefore are affine and relevant.

One issue with the above definition is that, just like for regular premonoidal
categories, it is difficult to distinguish between those morphisms that are
"truly" central versus those which are just central "by accident", as we only
want to require any notion of functor we develop subsequently to preserve the
former. We can solve this the same way, by simply extending the notion of
effectful category appropriately as follows:
\begin{definition}[Susbtructural Effectful Category]
  A \textbf{substructural effectful category} \(\mc{C}\) consists of
  \begin{itemize}
    \item A substructural monoidal category \(\mc{C}_1\); we call the morphisms
    in this category \textbf{central} or \textbf{linear}
    \item A substructural premonoidal category \(\mc{C}_0\) sharing the same
    affine and relevant objects as \(\mc{C}_1\)
    \item An identity-on-objects functor \(\upg{\cdot}{}: \mc{C}_1^\varnothing
    \to \mc{C}_0^\varnothing\) preserving all substructural premonoidal
    structure, i.e., preserves all symmetries, associators, unitors, drops, and
    splits and sends \(\mc{C}_1^q\) to \(\mc{C}_0^q\). In particular,
    restricting \(\upg{\cdot}{}\) appropriately makes all \(\mc{C}_1^q \to
    \mc{C}_0^q\) into effectful categories.
  \end{itemize}
  We define \(\upg{f}{0}\) to be \(\upg{f}{}\) for morphisms in \(\mc{C}_1^q\)
  and simply \(f\) for morphisms in \(\mc{C}_0\), and \(\upg{f}{1}\) to simply
  be the identity functor on \(\mc{C}_1^q\).

  Morphisms in \(\mc{C}_p^{\ms{r}}\) are called \textbf{copyable}, morphisms
  in \(\mc{C}_p^{\ms{a}}\) are called \textbf{discardable}, while morphisms
  in \(\mc{C}_1^{\ms{a}}, \mc{C}_1^{\ms{r}}, \mc{C}_1^{\{\ms{a},
  \ms{r}\}}\) specifically are called \textbf{affine}, \textbf{relevant}, and
  \textbf{intuitionistic} or \textbf{pure} respectively.
\end{definition}
We can obtain an example of a substructural effectful category quite easily by,
considering the power-set monad above, applying the state transformer to yield
the Kliesli category with morphisms of the form \(\mc{C}_0(A, B) = A \to S \to
\mc{P}(A \times S)\). Letting \(\mc{C}_1^q = \ms{Rel}^q\) and defining
\(\upg{R}{} = \lambda a, s. R(a) \times \{s\}\), we can verify that this
satisfies the axioms of a substructural effectful category. Note that here the
\textit{central} morphisms are precisely those which do not read from or write
to the state, and therefore which commute with all other morphisms.

@@

@@

Now that we have a semantics for \isotopessa programs, we can show that it
matches the syntax by showing that the syntactic and semantic notions of
substitution coincide. We begin by giving a semantics for the judgement
\(\issubst{\gamma}{\Theta}{\Gamma}\), which we will interpret as a pure morphism
taking \(\dnt{\Theta}\) to \(\dnt{\Gamma}\), defined via the rules in Figure
\ref{fig:ssa-subst-semantics}, which can simply be viewed as computing all
\(\gamma(x)\) for \(\thyp{x}{A}{q}\) in \(\Gamma\). In particular, we interpret
\rle{subst-nil} as the identity function, and \rle{subst-cons} by splitting the
context via \(\dnt{\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}}\), and then
passing \(\dnt{\Theta_\Gamma}\) to the remainder of the substitution
\(\dnt{\issubst{\gamma}{\Gamma}{\Theta_\Gamma}}\) and \(\dnt{\Theta_x}\) through
\(\dnt{\hasty{\Theta_x}{1}{\gamma(x)}{A}{q}}\) to yield \(\dnt{A}\). We give a
similar semantics for label-substitutions: \rle{lb-subst-nil} is interpreted as
the zero morphism \(0_{\dnt{\ms{K}}}: \mc{C}_0^\varnothing(\mb{0},
\dnt{\ms{K}})\), while \rle{lb-subst-cons} takes in the coproduct \(\dnt{\ms{L}}
+ \dnt{\Gamma} \otimes \dnt{A}\), sending the former to
\(\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{A}}}\) and the latter to the denotation of
\(\mc{L}(\lbl{\ell}, x)\).

\begin{figure}
  \begin{gather*}
    \boxed{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}: \mc{C}_1^\varnothing(\dnt{\Theta}, \dnt{\Gamma})} 
    \\
    \dnt{\issubst{\cdot}{\cdot}{\cdot}} = \ms{id}
    \qquad
    \dnt{\issubst{\gamma}{\Theta}{\Gamma, \thyp{x}{A}{q}}}
      = \dnt{\csplits{\Theta}{\Theta_\Gamma}{\Theta_x}};\dnt{\issubst{\gamma}{\Theta_\Gamma}{\Gamma}} \otimes \dnt{\hasty{\Theta_x}{1}{a}{A}{q}}
    \\
    \boxed{\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}: \mc{C}_0^\varnothing(\dnt{\ms{L}}, \dnt{\ms{K}})} 
    \\
    \dnt{\lbsubst{\mc{L}}{\cdot}{\ms{K}}} = 0_{\ms{K}}
    \qquad
    \dnt{\lbsubst{\mc{L}}{\ms{L}, \llhyp{\ell}{\Gamma}{A}}{\ms{K}}}
    = [\dnt{\lbsubst{\ms{L}}{\ms{L}}{\ms{K}}}, \dnt{\haslb{\Gamma, \thyp{x}{A}{}}{\mc{L}(\lbl{\ell}, x)}{\ms{K}}}]
  \end{gather*}
  \caption{Semantics for \isotopessa substitutions}
  \Description{Semantics for isotope-SSA substitutions}
  \label{fig:ssa-subst-semantics}
\end{figure}
